The locally Gaussian partial correlation

### 2021-01-22 lgpc_JBES.tex
Paper akseptert, rettet opp i noen små detaljer iht kommentar fra referee

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
\begin{document}
\maketitle
\begin{abstract}
It is well known in econometrics and other fields that the dependence structure for jointly Gaussian variables can be fully captured using correlations, and that the \textit{conditional} dependence structure in the same way can be described using partial correlations. The partial correlation does not, however, characterize conditional dependence in many non-Gaussian populations. This paper introduces the local Gaussian partial correlation (LGPC), a new measure of conditional dependence. It is a local version of the partial correlation coefficient that characterizes conditional dependence in a large class of populations. It has some useful and novel properties besides: The LGPC reduces to the ordinary partial correlation for jointly normal variables, and it distinguishes between positive and negative conditional dependence. Furthermore, the LGPC can be used to study departures from conditional independence in specific parts of the distribution. We provide several examples of this, both simulated and real, and derive estimation theory under a local likelihood estimation framework. Finally, we indicate how the LGPC can be used to construct a powerful test for conditional independence, which, for example, can be used to detect nonlinear Granger causality in time series. \newline \textbf{Keywords:} \emph{Conditional dependence, conditional independence test, Granger causality}
\end{abstract}

@@

In all of the following we assume that the components of \(\X\) are continuous with joint density function \(f_{\X}\). Given a point \(\x\), we approximate \(f_{\X}\) in a neighborhood of \(\x\) by a multivariate Gaussian density

\begin{equation} 
\psi(\x, \fv) = \frac{1}{(2\pi)^{p/2}|\fSigma(\x)|^{1/2}} \exp \left\{-\frac{1}{2}(\fv - \fmu(\x))^T\fSigma^{-1}(\x)(\fv - \fmu(\x))\right\},
\label{eq:localgaussian}
\end{equation}
where \(\x = (x_1\ldots, x_p)\), \(\fmu(\x) = \{\mu_j(\x)\}\) and \(\fSigma(\x) = \{\sigma_{jk}(\x)\}\) for \(j,k=1,\ldots, p\). Moving to another point \(\y\), there is another (generally different) Gaussian approximation \(\psi(\y, \fv)\). In this way we approximate \(f_{\X}\) by a family of multivariate Gaussian densities defined by a set of smooth parameter functions \(\{\fmu(\x), \fSigma(\x)\}\), and if \(f_{\X}\) is itself a Gaussian density, then the parameter functions collapse to constants corresponding to the true parameter values, and \(\psi(\x) \equiv f_{\X}(\x)\). Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}) provide the general framework for estimating such parameter functions non-parametrically from a given data set using a local likelihood procedure, and the basic idea in the following treatment is to replace the components in the partial covariance matrix \eqref{eq:matrixpartial} by their locally estimated counterparts in order to obtain a local measure of conditional dependence.

@@

In practice, we do not know \(F_{j}\), but we can instead use the empirical distribution function
\[\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),\]
where \(1(\cdot)\) is the indicator function, \(n\) is the number of observations, and \(X_{ji}\) is the \(i\)th observation of \(X_j\). This results in pseudo standard normal variables \(\widehat Z_{j} = \Phi^{-1}(\widehat F_{j}(X_j))\).

In the following, we will not always distinguish between \(Z_j\) and \(\widehat Z_{j}\). In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(\widehat Z_{j}\), instead of \(Z_j\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.

@@

\[f_{z_1|\z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},\]
where

\begin{equation}
\mu_{1|3}(z_1) = \R_{12}(z_1, \z_3)\R_{22}(\z_3)^{-1}\z_3 \,\,\, \textrm{ and } \,\,\, \sigma_{1|3}(z_1) = \fSigma_{Z_1|\Z_3},
\label{eq:localconditionalpar}
\end{equation}
and where the latter expression is defined analogously to \eqref{eq:matrixlocalpartial}. The conditional density \(f_{Z_2|\Z_3}(z_2|\z_3)\) is defined in the same way. If \(Z_1\) and \(Z_2\) are conditionally independent given \(\Z_3\), then

\begin{align*}
f_{Z_1,Z_2|\Z_3}(z_1, z_2|\z_3) &= f_{Z_1|\Z_3}(z_1|\z_3)f_{Z_2|\Z_3}(z_2|\z_3) \\
&= \frac{1}{2\pi\sqrt{1-\sigma_{1|3}(z_1)}\sqrt{1-\sigma_{2|3}(z_2)}} \\ 
&\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2)}{\sigma_{2|3}(z_2)}\right)^2\right\},
\end{align*}
and we at once identify the conditional density of \((Z_1,Z_2)|\Z_3\) as another Gaussian distribution, but without any cross-term involving \(z_1z_2\), which then implies that the off-diagonal element in the partial covariance matrix \eqref{eq:matrixlocalpartial} is identically equal to zero. We see immediately from our definition \eqref{eq:definition} that the LGPC is also identically equal to zero.

@@

\begin{equation}
\hfSigma_{12|3}(\z) = \hR^{11}(\z) -  \hR^{12}(\z)\left(\hR^{22}(\z)\right)^{-1}\hR^{21}(\z),
\label{eq:localpartialcov}
\end{equation}
and if we assume that \(X_1\) and \(X_2\) constitute the two first elements in \(\X\), then, following \eqref{eq:definition}, the local Gaussian partial correlation between \(X_1\) and \(X_2\) given \(\Xtwo = \X_3\) on the \(z\)-scale is estimated by
\begin{equation}
\widehat\alpha(\z) = \frac{\Big\{\hfSigma_{12|3}(\z)\Big\}_{12}}{\Big\{\hfSigma_{12|3}(\z)\Big\}_{11}^{1/2}\Big\{\hfSigma_{12|3}(\z)\Big\}_{22}^{1/2}}.
\label{eq:estimate}
\end{equation}

@@

We transform the observations to the \(z\)-scale as in \eqref{eq:pseudoobservations}, but now estimate the individual components in \(\R(\z)\) one by one, taking only into consideration the \emph{pair} of variables in question:
\begin{equation}
\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
\label{eq:locallik2}
\end{equation}
where all running variables and samples are bivariate subsets corresponding to the indices \((j,k)\), \(\psi(\cdot, \rho)\) is the bivariate version of \eqref{eq:lgdeapprox}, and \(\hh\) now is a \(2\times2\) diagonal matrix of bandwidths. After estimating all local correlations in this way, we proceed to calculate the LGPC using equations \eqref{eq:localpartialcov} and \eqref{eq:estimate} as above.

@@

Consider first the full trivariate fit \(\hR(\z)\) described in Section \ref{chap:trivariate-full} with \(\z = (z_1,z_2,z_3)\). We estimate the local correlation matrix \(\R(\z)\) by maximizing the local likelihood function in \eqref{eq:locallik3}, which we denote by \(L_n(\R(\z), \z)\), with the only exception that we, for now, assume the true transformation between the \(x\)- and the \(z\)-scale to be known. For a fixed matrix of bandwidths \(\hh\), denote by \(\R_{\hh}(z)\) the local correlation that satisfies, as \(n \rightarrow \infty\),

\begin{equation}
\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
\label{eq:popfixedh3}
\end{equation}
where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\) (at the point \(\z\)), and \(\fu(\cdot, \rho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \rho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\). Denote by \(\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}\) the vector of local correlations. The joint limiting distribution of \(\hfrho\) (for convenience only stated on the \(z\)-scale below) is given by the following result as \(n \rightarrow \infty\) and \(\hh \rightarrow \bm{0}\):

\BeginKnitrBlock{theorem}
\protect\hypertarget{thm:loccor3}{}{\label{thm:loccor3} }Let \(\{\Z_t\}_{t=1}^n\) be observations on the marginally standard normally distributed random vector \(\Z\) having joint density \(f_{\Z}(\z)\). Assume that the following conditions hold:

@@

Then, by writing \(b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0\) assuming they all converge to zero at the same rate,
\begin{equation}
\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
\label{eq:an3}
\end{equation}
where \(\bm{I}\) is the \(3\times3\) identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
\EndKnitrBlock{theorem}

@@

\BeginKnitrBlock{theorem}
\protect\hypertarget{thm:pseudo}{}{\label{thm:pseudo} }Under the conditions in Theorems \ref{thm:loccor3} and Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), assume further that the kernel function \(K(\cdot)\) has bounded support. Replacing the marginally standard normal vector \(\Z\) with the approximately marginally standard normal vector \(\hZ\), does not change the conclusions in these theorems.
\EndKnitrBlock{theorem}

\BeginKnitrBlock{proof}
\iffalse{} {Proof. } \fi{}See Section D of the online supplement.
\EndKnitrBlock{proof}

@@

\begin{equation}
R_{1,t} \perp R_{2, t-1} \,\,|\,\, R_{1,t-1},
\end{equation}
which we may test formally using the approaches described in the following section or the references therein. Let us rather concentrate on \emph{describing} this conditional dependence relationship by estimating the local Gaussian partial correlation between \(R_{1,t}\) and \(R_{2,t-1}\) given a few different values of \(R_{2,t-1}\), which gives considerable more information than a single \(p\)-value of a test. We collect daily observations on Oslo Stock Exchange\footnote{Thanks to \emph{Børsprosjektet (The Stock Exchange Project)} at the Norwegian School of Economics for providing the data.} from 1 January 2018 until 31 December 2019. In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal \(R_{1,t} = R_{2,t-1}\) for three different values \(R_{1,t-1} = -0.02, 0\) and \(0.02\), along with their block-bootstrapped 95\% confidence intervals. The constant \(\alpha(\z) = 0\) is indicated as a dotted line.

@@

A test based on the LGPC is quite different from the methods quoted above. It is semi-parametric and does not rely on traditional kernel estimates of density-, \mbox{distribution-,} or characteristic functions. Furthermore, due to our transformation of the data \eqref{eq:trans} to marginal standard normality, we do not necessarily have to specify weight functions in our test functional to lessen the impact of outliers - unless of course we wish to test for conditional independence in a specific portion of the sample space.

On the other hand, we have to pay a price when imposing structure on the dependence. Certain types of conditional dependence will remain invisible to our test statistic, but simulation experiments show that our test performs on par with, and sometimes better than existing, fully nonparametric tests.

@@

\begin{equation}
T_{n,b} = \int_S h\left( \widehat \alpha_b(\z)  \right) \di F_n(\z),
\label{eq:teststatistic}
\end{equation}
where \(h(\cdot)\) is a real-valued function that is typically even and non-negative for most standard applications, and \(S \subseteq \mathbb{R}^p\) is an integration area that can be altered in order to focus the test to specific portions of the sample space. Standard laws of large numbers ensure that, under regularity conditions, see Theorem F.1 of the supplement, \(T_{n,b}\) converges in probability towards its population value \[T = \int_S h\left(\alpha(\z)\right) \di F(\z).\]

>>>>

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle
\begin{abstract}
It is well known in econometrics and other fields that the dependence structure for jointly Gaussian variables can be fully captured using correlations, and that the \textit{conditional} dependence structure in the same way can be described using partial correlations. The partial correlation does not, however, characterize conditional dependence in many non-Gaussian populations. This paper introduces the local Gaussian partial correlation (LGPC), a new measure of conditional dependence. It is a local version of the partial correlation coefficient that characterizes conditional dependence in a large class of populations. It has some useful and novel properties besides: The LGPC reduces to the ordinary partial correlation for jointly normal variables, and it distinguishes between positive and negative conditional dependence. Furthermore, the LGPC can be used to study departures from conditional independence in specific parts of the distribution. We provide several examples of this, both simulated and real, and derive estimation theory under a local likelihood estimation framework. Finally, we indicate how the LGPC can be used to construct a powerful test for conditional independence, which, for example, can be used to detect nonlinear Granger causality in time series. \newline \textbf{Keywords:} \emph{Conditional dependence, conditional independence test, Granger causality}
\end{abstract}

@@

In all of the following we assume that the components of \(\X\) are continuous with joint density function \(f_{\X}\). Given a point \(\x\), we approximate \(f_{\X}\) in a neighborhood of \(\x\) by a multivariate Gaussian density \(\psi({\x},{\fv})\), which at the point \({\fv} = {\x}\) is

\begin{equation} 
\psi(\x) = \frac{1}{(2\pi)^{p/2}|\fSigma(\x)|^{1/2}} \exp \left\{-\frac{1}{2}(\x - \fmu(\x))^T\fSigma^{-1}(\x)(\x - \fmu(\x))\right\},
\label{eq:localgaussian}
\end{equation}
where \(\x = (x_1\ldots, x_p)\), \(\fmu(\x) = \{\mu_j(\x)\}\) and \(\fSigma(\x) = \{\sigma_{jk}(\x)\}\) for \(j,k=1,\ldots, p\). Moving to another point \(\y\), there is another (generally different) Gaussian approximation \(\psi(\y, \fv)\). In this way we approximate \(f_{\X}\) by a family of multivariate Gaussian densities defined by a set of smooth parameter functions \(\{\fmu(\x), \fSigma(\x)\}\), and if \(f_{\X}\) is itself a Gaussian density, then the parameter functions collapse to constants corresponding to the true parameter values, and \(\psi(\x) \equiv f_{\X}(\x)\). Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}) provide the general framework for estimating such parameter functions non-parametrically from a given data set using a local likelihood procedure, and the basic idea in the following treatment is to replace the components in the partial covariance matrix \eqref{eq:matrixpartial} by their locally estimated counterparts in order to obtain a local measure of conditional dependence.

@@

In practice, we do not know \(F_{j}\), but we can instead use the empirical distribution function
\[\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{j,t} \leq x\right),\]
where \(1(\cdot)\) is the indicator function, \(n\) is the number of observations, and \(X_{j,t}\) is the \(t\)th observation of \(X_{j}\). This results in pseudo standard normal variables \(\widehat Z_{j,t} = \Phi^{-1}(\widehat F_{j}(X_{j,t}))\).

In the following, we will not always distinguish between \(Z_{j,t}\) and \(\widehat Z_{j,t}\). In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(\widehat Z_{j,t}\), instead of \(Z_{j,t}\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.

@@

\[f_{Z_1|\Z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1, \z_3)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1, \z_3)}{\sigma_{1|3}(z_1, \z_3)}\right)^2\right\},\]
where

\begin{equation}
\mu_{1|3}(z_1, \z_3) = \R_{12}(z_1, \z_3)\R_{22}(\z_3)^{-1}\z_3 \,\,\, \textrm{ and } \,\,\, \sigma_{1|3}(z_1, \z_3) = \fSigma_{Z_1|\Z_3}(z_1,\z_3),
\label{eq:localconditionalpar}
\end{equation}
and where the latter expression is defined analogously to \eqref{eq:matrixlocalpartial}. The conditional density \(f_{Z_2|\Z_3}(z_2|\z_3)\) is defined in the same way. If \(Z_1\) and \(Z_2\) are conditionally independent given \(\Z_3\), then, by definition,

\begin{align*}
f_{Z_1,Z_2|\Z_3}(z_1, z_2|\z_3) &= f_{Z_1|\Z_3}(z_1|\z_3)f_{Z_2|\Z_3}(z_2|\z_3) \\
&= \frac{1}{2\pi\sqrt{1-\sigma_{1|3}(z_1, \z_3)}\sqrt{1-\sigma_{2|3}(z_2, \z_3)}} \\ 
&\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1, \z_3)}{\sigma_{1|3}(z_1, \z_3)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2, \z_3)}{\sigma_{2|3}(z_2, \z_3)}\right)^2\right\},
\end{align*}
and we at once identify the conditional density of \((Z_1,Z_2)|\Z_3\) as another Gaussian distribution, but without any cross-term involving \(z_1z_2\), which then implies that the off-diagonal element in the partial covariance matrix \eqref{eq:matrixlocalpartial} is identically equal to zero. We see immediately from our definition \eqref{eq:definition} that the LGPC is also identically equal to zero.

@@

\begin{equation}
\hfSigma_{12|3}(\z) = \hR^{11}(\zone) -  \hR^{12}(\z)\left(\hR^{22}(\ztwo)\right)^{-1}\hR^{21}(\z),
\label{eq:localpartialcov}
\end{equation}
and if we assume that \(X_1\) and \(X_2\) constitute the two first elements in \(\X\), then, following \eqref{eq:definition}, the local Gaussian partial correlation between \(X_1\) and \(X_2\) given \(\Xtwo = \X_3\) on the \(z\)-scale is estimated by
\begin{equation}
\widehat\alpha(\z) = \frac{\Big\{\hfSigma_{12|3}(\z)\Big\}_{12}}{\Big\{\hfSigma_{12|3}(\z)\Big\}_{11}^{1/2}\Big\{\hfSigma_{12|3}(\z)\Big\}_{22}^{1/2}}.
\label{eq:estimate}
\end{equation}

@@

We transform the observations to the \(z\)-scale as in \eqref{eq:pseudoobservations}, but now estimate the individual components in \(\R(\z)\) one by one, taking only into consideration the \emph{pair} of variables in question:
\begin{align}
\hrho_{jk}(z_j, z_k) &= \arg\max_{\rho_{jk}} \Bigg[\frac{1}{n} \sum_{t=1}^nK_b(\widehat{Z}_{j,t} - z_j, \widehat{Z}_{k,t}- z_k) \log(\psi(\widehat{Z}_{j,t}, \widehat{Z}_{k,t}; \rho_{jk}(z_j, z_k))) \nonumber \\
& \qquad\qquad\qquad\qquad\qquad - \int K_b(v - z_j, w- z_k)\psi(v,w; \rho_{jk}(z_j,z_k))\di v \di w,
\label{eq:locallik2}
\end{align}
where all running variables and samples are bivariate subsets corresponding to the indices \((j,k)\), \(\psi(\cdot, \rho)\) is the bivariate version of \eqref{eq:lgdeapprox}, and \(\hh\) now is a \(2\times2\) diagonal matrix of bandwidths. After estimating all local correlations in this way, we proceed to calculate the LGPC using equations \eqref{eq:localpartialcov} and \eqref{eq:estimate} as above.

@@

Consider first the full trivariate fit \(\hR(\z)\) described in Section \ref{chap:trivariate-full} with \(\z = (z_1,z_2,z_3)\). We estimate the local correlation matrix \(\R(\z)\) by maximizing the local likelihood function in \eqref{eq:locallik3}, which we denote by \(L_n(\R(\z), \z)\), with the only exception that we, for now, assume the true transformation between the \(x\)- and the \(z\)-scale to be known. For a fixed matrix of bandwidths \(\hh\), denote by \(\frho_{\hh}(\z) = \{\rho_{12,\hh}(\z), \rho_{13,\hh}(\z), \rho_{23,\hh}(\z)\}\) the vector of local correlations that satisfiy, as \(n \rightarrow \infty\),

\begin{equation}
\nabla L_n(\frho_{\hh}(\z), \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \frho_{\hh})\}\,\textrm{d}\y = 0,
\label{eq:popfixedh3}
\end{equation}
where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\) (at the point \(\z\)), and \(\fu(\cdot, \frho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \frho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\). The joint limiting distribution of \(\hfrho(\z)\) (for convenience only stated on the \(z\)-scale below) is given by the following result as \(n \rightarrow \infty\) and \(\hh \rightarrow \bm{0}\):

\begin{theorem}
\protect\hypertarget{thm:loccor3}{}{\label{thm:loccor3} }Let \(\{\Z_t\}_{t=1}^n\) be observations on the marginally standard normally distributed random vector \(\Z\) having joint density \(f_{\Z}(\z)\). Assume that the following conditions hold:

@@

Then, by writing \(b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0\) assuming they all converge to zero at the same rate,
\begin{equation}
\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
\label{eq:an3}
\end{equation}
where \(\bm{I}\) is the \(3\times3\) identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
\end{theorem}

@@

\begin{theorem}
\protect\hypertarget{thm:pseudo}{}{\label{thm:pseudo} }Under the conditions in Theorems \ref{thm:loccor3} and Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), assume further that the kernel function \(K(\cdot)\) has bounded support. Replacing the marginally standard normal vector \(\Z\) with the approximately marginally standard normal vector \(\hZ\), does not change the conclusions in these theorems.
\end{theorem}

\begin{proof}
\iffalse{} {Proof. } \fi{}See Section D of the online supplement.
\end{proof}

@@

\begin{equation}
R_{1,t} \perp R_{2, t-1} \,\,|\,\, R_{1,t-1},
\end{equation}
which we may test formally using the approaches described in the following section or the references therein. Let us rather concentrate on \emph{describing} this conditional dependence relationship by estimating the local Gaussian partial correlation between \(R_{1,t}\) and \(R_{2,t-1}\) given a few different values of \(R_{2,t-1}\), which gives considerable more information than a single \(p\)-value of a test. We collect daily observations on Oslo Stock Exchange\footnote{Thanks to \emph{Børsprosjektet (The Stock Exchange Project)} at the Norwegian School of Economics for providing the data.} from 1 January 2018 until 31 December 2019. In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal \(R_{1,t} = R_{2,t-1}\) for three different values \(R_{1,t-1} = -0.02, 0\) and \(0.02\), along with their block-bootstrapped 95\% confidence intervals. The constant \(\alpha(\z) = 0\) is indicated as a dotted line. A desirable property of the LGPC is that one is able to distinguish between positive and negative conditional dependence. In the context of Granger causality, we view this as a nice empirical device in contrast to the conventional Granger causality testing procedure that fails to say anything about the potential asymmetric pattern in the underlying problem.

@@

A test based on the LGPC is quite different from the methods quoted above. It is semi-parametric and does not rely on traditional kernel estimates of density-, \mbox{distribution-,} or characteristic functions. Furthermore, due to our transformation of the data \eqref{eq:trans} to marginal standard normality, we do not necessarily have to specify weight functions in our test functional to lessen the impact of outliers - unless of course we wish to test for conditional independence in a specific portion of the sample space. On the other hand, we have to pay a price when imposing structure on the dependence. Certain types of conditional dependence will remain invisible to our test statistic, but simulation experiments show that our test performs on par with, and sometimes better than existing, fully nonparametric tests.

@@

\begin{equation}
T_{n,b} = \int_S h\left( \widehat \alpha_b(\z)  \right) \di F_n(\z),
\label{eq:teststatistic}
\end{equation}
where \(h(\cdot)\) is a real-valued function that is typically even and non-negative for most standard applications (we use \(h(x) = x^2\) in all our examples), and \(S \subseteq \mathbb{R}^p\) is an integration area that can be altered in order to focus the test to specific portions of the sample space. Standard laws of large numbers ensure that, under regularity conditions, see Theorem F.1 of the supplement, \(T_{n,b}\) converges in probability towards its population value \[T = \int_S h\left(\alpha(\z)\right) \di F(\z).\]
